Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-24 ind_files/03_BD_processes.tex
Minor changes for reviewer comments

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{ For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
It is easy to see that if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. Similar well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can also be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

>>>>

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{ For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
In this equation, if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. To see this, one can observe that the RHS of equation \eqref{two_species_stoch_RM} identically vanishes at both $p=0$ and $p=1$, since every term on the RHS contains the product $p(1-p)$. Further, since there is no mutation between types, $p=0$ (extinction of type 1 organisms) and $p=1$ (extinction of type 2 organisms) are both `absorbing states', \emph{i.e.} a system which reaches these states can never leave them. Well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

### 2023-04-24 ind_files/06_nontechnical_discussion.tex
Minor changes for reviewer comments

Actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will likely quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}\textcolor{blue}{, and innovative future work is needed to develop tractable approximation schemes to handle these complications}. However, while such complicated behavior may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with. Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological populations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.

>>>>

Actually solving the equations I formulate analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will likely quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}\textcolor{blue}{, and innovative future work is needed to develop tractable approximation schemes to handle these complications}. However, while such complicated behavior may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with. Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological populations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.
